---
output: github_document
---

[![Travis-CI Build Status](https://travis-ci.org/codemeta/codemetar.svg?branch=master)](https://travis-ci.org/codemeta/codemetar)
[![Coverage Status](https://img.shields.io/codecov/c/github/codemeta/codemetar/master.svg)](https://codecov.io/github/codemeta/codemetar?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/codemetar)](https://cran.r-project.org/package=codemetar)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

# codemetar

The goal of codemetar is to generate the JSON-LD file, `codemeta.json` containing software metadata describing an R package

## Installation

You can install codemetar from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("codemeta/codemetar")
```

## Example

This is a basic example which shows you how to generate a `codemeta.json` for your R package:

```{r example}
codemetar::write_codemeta("testthat")
```

```{r include=FALSE}
unlink("codemeta.json")
```


Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

