---
title: "Intro to codemetar, Codemeta creator for R packages"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to the CodeMeta Project and codemetar package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r include=FALSE}
knitr::opts_chunk$set(comment="")
if(grepl("windows", tolower(Sys.info()[["sysname"]])))
  knitr::opts_chunk$set(comment="", error =TRUE)
```



# codemetar: generate codemeta metadata for R packages

```{r child='./rmdhunks/goals.Rmd'} 
```


## Motivation

```{r child='./rmdhunks/whybother.Rmd'} 
```


## Create a codemeta.json in one function call

```{r child='./rmdhunks/example.Rmd'} 
```


# A brief intro to common terms we'll use:

- [Linked data](https://en.wikipedia.org/wiki/Linked_data): We often use different words to mean the same thing. And sometimes the same word to mean different things.  Linked data seeks to address this issue by using [URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier) (i.e. URLs) to make this explcit.

- context: No one likes typing out long URLs all the time.  So instead, the *context* of a JSON-LD file (`"@context"` element) gives us the context for the terms we use, that is, the root URL.  Usually schema.org but domain specific ones also (eg codemeta)

- [Schema.org](http://schema.org): A major initiative led by Google and other search engines to define a simple and widely used context to link data on the web through a catalogue of standard metadata fields

- [The CodeMeta Project](http://codemeta.github.io): an academic led community initiative to formalise the metadata fields included in typical software metadata records and introduce important fields that did not have clear equivalents. The codemeta [crosswalk](https://codemeta.github.io/crosswalk) provides an explicit map between the metadata fields used by a broad range of software repositories, registries and archives

- [JSON-LD](https://json-ld.org): While 'linked data' can be represented in many different formats, these have consistently proven a bit tricky to use, either for consumers or developers or both.  JSON-LD provides a simple adaptation of the JSON format, which has proven much more popular with both audiences, that allows it to express (most) linked-data concepts.  It is now the format of choice for expressing linked data by Google and many others.  Any JSON-LD file is valid JSON, and any JSON file can be treated as JSON-LD.  

- [codemetar](https://codemeta.github.io/codemetar): The CodeMeta Project has created tools in several languages to impelement the CodeMeta Crosswalk (using JSON-LD) and help extract software metadata into `codemeta.json` records.  `codemetar` is one such tool, focused on R and R packages.

## Modifying or enriching CodeMeta metadata

The best way to ensure `codemeta.json` is as complete as possible is to begin by making full use of the fields that can be set in an R package DESCRIPTION file, such as `BugReports` and `URL`.  Using the `Authors@R` notation allows a much richer specification of author roles, correct parsing of given vs family names, and email addresses. 

In the current implementation, developers may specify an ORCID url for an author in the optional `comment` field of `Authors@R`, e.g.

```
Authors@R: person("Carl", "Boettiger", role=c("aut", "cre", "cph"), email="cboettig@gmail.com", comment="http://orcid.org/0000-0002-1642-628X")
```

which will allow `codemetar` to associate an identifier with the person.  This is clearly something of a hack since R's `person` object lacks an explicit notion of `id`, and may be frowned upon.  






### Using the DESCRIPTION file

The DESCRIPTION file is the natural place to specify any metadata for an R package.  The `codemetar` package can detect certain additional terms in the [CodeMeta context](https://codemeta.github.io/terms).  Almost any additional codemeta field can be added to and read from the DESCRIPTION into a `codemeta.json` file (see `codemetar:::additional_codemeta_terms` for a list).  

CRAN requires that you prefix any additional such terms to indicate the use of `schema.org` explicitly, e.g. `keywords` would be specified in a DESCRIPTION file as:

```
X-schema.org-keywords: metadata, codemeta, ropensci, citation, credit, linked-data
```




Where applicable, these will override values otherwise guessed from the source repository.  Use comma-separated lists to separate multiple values to a property, e.g. keywords.  

See the [DESCRIPTION](https://github.com/codemeta/codemetar/blob/master/DESCRIPTION) file of the `codemetar` package for an example.  

```{r include=FALSE}
#unlink("codemeta.json")
```


## Going further

Check out all the [codemetar vignettes](https://codemeta.github.io/codemetar/articles/index.html) for tutorials on other cool stuff you can do with codemeta and json-ld.  

