---
title: "Intro to codemetar, Codemeta creator for R packages"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to the CodeMeta Project and codemetar package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r include=FALSE}
knitr::opts_chunk$set(comment="")
if(grepl("windows", tolower(Sys.info()[["sysname"]])))
  knitr::opts_chunk$set(comment="", error =TRUE)
```



# codemetar: generate codemeta metadata for R packages

```{r child='./rmdhunks/goals.Rmd'} 
```


## Motivation

```{r child='./rmdhunks/whybother.Rmd'} 
```


## Create a codemeta.json in one function call

```{r child='./rmdhunks/example.Rmd'} 
```


## A brief intro to common terms we'll use:

- [Linked data](https://en.wikipedia.org/wiki/Linked_data): We often use different words to mean the same thing. And sometimes the same word to mean different things.  Linked data seeks to address this issue by using [URIs](https://en.wikipedia.org/wiki/Uniform_Resource_Identifier) (i.e. URLs) to make this explcit.

- context: No one likes typing out long URLs all the time.  So instead, the *context* of a JSON-LD file (`"@context"` element) gives us the context for the terms we use, that is, the root URL.  Usually schema.org but domain specific ones also (eg codemeta)

- [Schema.org](http://schema.org): A major initiative led by Google and other search engines to define a simple and widely used context to link data on the web through a catalogue of standard metadata fields

- [The CodeMeta Project](http://codemeta.github.io): an academic led community initiative to formalise the metadata fields included in typical software metadata records and introduce important fields that did not have clear equivalents. The codemeta [crosswalk](https://codemeta.github.io/crosswalk) provides an explicit map between the metadata fields used by a broad range of software repositories, registries and archives

- [JSON-LD](https://json-ld.org): While 'linked data' can be represented in many different formats, these have consistently proven a bit tricky to use, either for consumers or developers or both.  JSON-LD provides a simple adaptation of the JSON format, which has proven much more popular with both audiences, that allows it to express (most) linked-data concepts.  It is now the format of choice for expressing linked data by Google and many others.  Any JSON-LD file is valid JSON, and any JSON file can be treated as JSON-LD.  

- [codemetar](https://codemeta.github.io/codemetar): The CodeMeta Project has created tools in several languages to impelement the CodeMeta Crosswalk (using JSON-LD) and help extract software metadata into `codemeta.json` records.  `codemetar` is one such tool, focused on R and R packages.

## Modifying or enriching CodeMeta metadata

```{r child='./rmdhunks/improvecodemetadata.md'} 
```

## Going further

Check out all the [codemetar vignettes](https://codemeta.github.io/codemetar/articles/index.html) for tutorials on other cool stuff you can do with codemeta and json-ld.  

