---
title: "Parsing Codmeta Data"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrate the kind of information we can discover by parsing collections of codemeta documents programmatically.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r include=FALSE}
knitr::opts_chunk$set(comment="")
```


Here we illustrate some example use cases that involve parsing codemeta data.  

```{r message=FALSE}
library(jsonld)
library(jsonlite)
library(magrittr)
library(codemetar)
library(purrr)
library(dplyr)
library(printr)
```

We start with a simple example from the `codemeta.json` file of `codemetar` itself.  First, we'll just generate a copy of the codemeta record for the package:

```{r}
write_codemeta("codemetar", "codemeta.json")
```

We then digest thus input using a JSON-LD "frame." While not strictly necessary, this helps ensure the data matches the format we expect, even if the original file had errors or missing data.  See the vignette "Validating in JSON-LD" in this package and the official [JSON-LD docs](https://json-ld.org/spec/latest/json-ld-framing/) for details).  The `codemetar` package includes a reasonably explicit frame to get us started:

```{r}
frame <- system.file("schema/frame_schema.json", package="codemetar")

meta <- 
  jsonld_frame("codemeta.json", frame) %>%
  fromJSON(FALSE) %>% getElement("@graph") %>% getElement(1)
```

Construct a citation

```{r}
authors <- 
lapply(meta$author, 
       function(author) 
         person(given = author$given, 
                family = author$family, 
                email = author$email,
                role = "aut"))
year <- meta$datePublished
if(is.null(year)) 
  year <- format(Sys.Date(), "%Y")
bibitem <- 
 bibentry(
     bibtype = "Manual",
     title = meta$name,
     author = authors,
     year = year,
     note = paste0("R package version ", meta$version),
     url = meta$URL,
     key = meta$identifier
   )

cat(format(bibitem, "bibtex"))

bibitem
```


## Parsing the ropensci corpus

The ropensci corpus consists of a list of codemeta files for all packages provided by the rOpenSci project, <ropensci.org>.  This provides a good test-case for how a large collection of codemeta files can be manipulated to help us get a better picture of the corpus.  

```{r}
download.file("https://github.com/codemeta/codemetar/raw/master/inst/notebook/ropensci.json",
              "ropensci.json")
```


As before, it is helpful, though not essential, to start off by framing the input data.

```{r}
frame <- system.file("schema/frame_schema.json", package="codemetar")

corpus <- 
    jsonld_frame("ropensci.json", frame) %>%
    fromJSON(simplifyVector = FALSE) %>%
    getElement("@graph") 
```

We're now ready to start exploring.  As usual, functions from `purrr` prove very useful for iterating through large JSON files.  First, we look at some basic summary data:

```{r}
## deal with nulls explicitly by starting with map
pkgs <- map(corpus, "name") %>% compact() %>% as.character()

# keep only those with package identifiers (names)
keep <- map_lgl(corpus, ~ length(.x$identifier) > 0)
corpus <- corpus[keep]

## now we can just do
all_pkgs <- map_chr(corpus, "name")
head(all_pkgs)
```

```{r}
## 60 unique maintainers
map_chr(corpus, c("maintainer", "familyName")) %>% unique() %>% length()

## Mostly Scott
map_chr(corpus, c("maintainer", "familyName")) %>% 
  as_tibble() %>%
  group_by(value) %>%
  tally(sort=TRUE)
```


```{r}
## number of co-authors ... 
map_int(corpus, function(r) length(r$author)) %>% 
  as_tibble() %>%
  group_by(value) %>%
  tally(sort=TRUE)
```

```{r}
## Contributors isn't used as much...
map_int(corpus, function(r) length(r$contributor)) %>% 
  as_tibble() %>%
  group_by(value) %>%
  tally(sort=TRUE)

```

Numbers (n) of packages with a total of (value) dependencies:

```{r}
map_int(corpus, function(r) length(r$softwareRequirements))  %>% 
  as_tibble() %>%
  group_by(value) %>%
  tally(sort=TRUE)
```

which dependencies are used most frequently?

```{r}
corpus %>%
map_df(function(x){
  ## single, unboxed dep
  if("name" %in% names(x$softwareRequirements))
    dep <- x$name
  else if("name" %in% names(x$softwareRequirements[[1]]))
    dep <- map_chr(x$softwareRequirements, "name")
  else { ## No requirementsß
    dep <- NA
  }
  
  tibble(identifier = x$identifier, dep = dep)
}) -> dep_df


dep_df %>%
group_by(dep) %>% 
  tally(sort = TRUE)

```

Alternate approach using a frame instead of `purrr` functions for subsetting the data.  Note that this gets all Depends and suggests (really all `SoftwareApplication` types mentioned)

```{r}
dep_frame <- '{
  "@context": "https://raw.githubusercontent.com/codemeta/codemeta/master/codemeta.jsonld",
  "@explicit": "true",
  "name": {}
}'
jsonld_frame("ropensci.json", dep_frame) %>% 
  fromJSON() %>% 
  getElement("@graph") %>%
  filter(type == "SoftwareApplication") %>%
  group_by(name) %>% 
  tally(sort = TRUE)
  
#  summarise(count(name))
```

```{r include = FALSE}
unlink("ropensci.json")
unlink("codemeta.json")
```
