---
title: "Translating between schema using JSON-LD"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translating between schema using JSON LD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r include=FALSE}
knitr::opts_chunk$set(comment="")
```

```{r message=FALSE}
library("codemetar")
library("tidyverse")
library("jsonlite")
library("jsonld")
library("httr")
```

One of the central motivations of JSON-LD is making it easy to translate between different representations of what are fundamentally the same data types. Doing so uses the two core algorithms of JSON-LD: *expansion* and *compaction*, as [this excellent short video by JSON-LD creator Manu Sporny](https://www.youtube.com/watch?v=Tm3fD89dqRE) describes.

Here's how we would use JSON-LD (from R) to translate between the two examples of JSON data from different providers as shown in the video.  First, the JSON from the original provider:

```{r}
ex <-
'{
"@context":{
  "shouter": "http://schema.org/name",
  "txt": "http://schema.org/commentText"
},
"shouter": "Jim",
"txt": "Hello World!"
}'
```

Next, we need the context of the second data provider.  This will let us translate the JSON format used by provider one ("Shouttr") to the second ("BigHash"):

```{r}
bighash_context <- 
'{
"@context":{
  "user": "http://schema.org/name",
  "comment": "http://schema.org/commentText"
}
}'
```

With this in place, we simply expand the original JSON and then compact using the new context:

```{r}
jsonld_expand(ex) %>%
  jsonld_compact(context = bighash_context)
```

## Crosswalk contexts

The CodeMeta Crosswalk table seeks to accomplish a very similar goal.  The crosswalk table provides a human-readable mapping of different software metadata providers into the codemeta context (an extension of schema.org).  

We'll start with the actual crosswalk table itself:

```{r message=FALSE}
crosswalk <- "https://github.com/codemeta/codemeta/raw/master/crosswalk.csv"
cw <- read_csv(crosswalk)
```


## DataCite


We can illustrate using the expansion and compaction process to crosswalk DataCite's metadata schema into other formats, such as the native codemeta format or the Zenodo format.  First, let's start with some DataCite metadata:


```{r include=FALSE, eval=FALSE}
## nope, sadly this is not quite clean json translation 

GET("https://doi.org/10.5281/zenodo.573741", 
    add_headers(Accept="application/vnd.datacite.datacite+xml")) %>% 
  content(type="application/xml") %>% 
  xml2::as_list() %>%  
  toJSON(pretty = TRUE, auto_unbox=TRUE)

## Read tidy version instead
```



```{r}
datacite_ex <- "https://raw.githubusercontent.com/codemeta/codemetar/master/inst/examples/datacite-xml.json"
cat(readLines(datacite_ex), sep="\n")
```

Note this example uses a JSON-ified representation of DataCite's internal schema, which is XML-based.  So far, this is just plain JSON, with no context.  We will use the crosswalk table to define the DataCite context with reference to codemeta:


```{r}
## read the file
datacite_list <- read_json(datacite_ex)

## add context using the codemeta crosswalk function on the current crosswalk table
datacite_list$`@context` <- codemetar::crosswalk("DataCite", cw)

## serialize as JSON
datacite_json <-
datacite_list %>% 
  toJSON(auto_unbox=TRUE, pretty=TRUE)

```



We can now expand this json in terms of DataCite context we just derived from the table, and then compact it into the CodeMeta native context:

```{r}
datacite_cm <-
  jsonld_expand(datacite_json) %>% 
  jsonld_compact(context = "https://raw.githubusercontent.com/codemeta/codemeta/master/codemeta-v2.jsonld") 
datacite_cm
```

The result should be a valid `codemeta.json` document. 


## Comparison to native schema.org translation:

DataCite actually returns data directly in JSON-LD format using exclusively the schema.org context.  We can query that data directly:

```{r}
library("httr")
resp <- GET("https://doi.org/10.5281/zenodo.573741", add_headers(Accept="application/vnd.schemaorg.ld+json"))
datacite_jsonld <- content(resp, type="application/json")
```

This time we have no need to invoke the crosswalk context, since the data is already in schema.org context (which codemeta also uses):

```{r}
datacite_cm <- 
  datacite_jsonld %>% 
  toJSON(pretty = TRUE, auto_unbox = TRUE) %>%
  jsonld_expand() %>% 
  jsonld_compact(context = "https://raw.githubusercontent.com/codemeta/codemeta/master/codemeta-v2.jsonld") 

datacite_cm
```


This is very close to the codemeta document we got by translating the DataCite XML using the crosswalk table.  


## Transforming into other column schema


Note that we can also use this approach to crosswalk into another standard by compacting into the context for the given column.    For instance, here we translate our datacite record into a zenodo record.  First we get the context from the crosswalk:

```{r}
zenodo_context <- 
  list("@context" = crosswalk("Zenodo", cw)) %>% 
  toJSON(pretty = TRUE, auto_unbox = TRUE)
```

then apply it as before:

```{r}
  jsonld_expand(datacite_cm) %>% 
  jsonld_compact(context = zenodo_context) 

```

This data should be now use valid Zenodo terms (e.g. using `creators` for `schema:author`).  Note that we also see explicitly the context that `codemeta::crosswalk` produced as `zenodo_context`, since we passed this information as a json file and not a URL.  This provides a nice illustration of what the crosswalk is returning when asked for a context of a given column.  

Note that data that many of the fields could not be directly crosswalked into Zenodo terms, because (at least according to the crosswalk table) they have no analog in Zenodo, such as `schema:dateCreated`.  Instead of dropping this data, the JSON-LD has retained the information using the original prefix `schema:`, indicating that the term is part of the `schema.org` context but not recognized in the context of Zenodo.  

Note this also happens if properties do not have a 1:1 map, since JSON-LD doesn't provide a convenient way to map two separate properties like `givenName`, `familyName` into a single property, `name` either.  JSON-LD will also refuse to map a term that has a potentially different type in the new context (since type coercion can also lead to data loss).  In this way, compaction is very conservative.  








## Codemeta

We can use a similar crosswalk strategy to translate from a JSON-LD file writing using v1 of the codemeta context into the current, v2 context:

```{r}
codemetav1 <- crosswalk("codemeta-V1", cw)

## Add some altered Type definitions from v1
codemetav1 <- c(codemetav1,
                    person = "schema:Person",
                    organization = "schema:Organization",
                    SoftwareSoureCode = "schema:SoftwareSourceCode")



```


To crosswalk a vocabulary, we must use the context the CrossWalk table defines for that vocabulary.  Most vocabularies are not already ontologies with their own base URLs, in which case this is relatively intuitive.  But we also need to use the CrossWalk context even when the data originates in another context (SoftwareOntology, DOAP, codemeta-v1), since otherwise those terms will stay in their original space (e.g `dcterms:identifer` will remain a different kind of property from a `schema:identifer`, etc)

```{r}
v1 <- "https://raw.githubusercontent.com/codemeta/codemetar/master/inst/examples/codemeta-v1.json"
#v1 <- "../examples/codemeta-v1.json" ## or local path
v1_obj <- jsonlite::read_json(v1)
v1_obj$`@context` <- codemetav1
v1_json <- toJSON(v1_obj, pretty=TRUE, auto_unbox = TRUE)

#write_json(v1_obj, "v1.json", pretty=TRUE, auto_unbox = TRUE)
```


With a JSON object in place, along with a context file that defines how that vocabulary is expressed in the context of `schema.org` / `codemeta`, we can now perform the exansion and compaction routine as before.  Expansion uses the native context from the file, compaction uses the new context of codemeta terms.  


```{r}
v2_json <- 
jsonld_expand(v1_json) %>%
  jsonld_compact(context = "https://raw.githubusercontent.com/codemeta/codemeta/master/codemeta-v2.jsonld") 
v2_json
```

Note that certain terms in the `codemeta` namespace are explicitly being typed as such (e.g. `codemeta:maintainer` rather than plain `maintainer`) by the compaction algorithm, because these terms do not have matching types in their original codemeta v1 context vs codemeta v2 context.  




## Framing


We can use a frame to extract particular elements in a particular format.  This is most useful when there are highly nested complex types.  Framing with the `@explicit` tag is also a good way to filter out fields that we are not interested in, though these are usually less problematic for developers to work around.

```{r}
frame <- 
'{
  "@context":"https://raw.githubusercontent.com/codemeta/codemeta/master/codemeta-v2.jsonld",
  "@type": "SoftwareSourceCode",
  "@explicit": "true",
  "readme": {},
  "description": {},
  "maintainer": {}
}'

jsonld_frame(v2_json, frame)

```


Note that our frame can refer to `maintainer` even though the compaction has left the Property as `codemeta:maintainer`.  





